---
title: "Washington ZEV Demand After Two Market Shocks"
subtitle: "Monthly title transactions, fuel pressure, and stated preferences"
author: "MSBA 580 Project Group"
date: "`r format(Sys.Date(), '%B %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    self_contained: true
---

```{r setup, include=FALSE}
report_path <- normalizePath(knitr::current_input(dir = TRUE), mustWork = TRUE)
project_root <- normalizePath(file.path(dirname(report_path), ".."), mustWork = TRUE)
for (script_name in c("run_washington_analysis.R", "run_conjoint_analysis.R")) {
  status <- system2(
    file.path(R.home("bin"), "Rscript"),
    file.path(project_root, "scripts", script_name)
  )
  if (status != 0L) stop(script_name, " did not complete.")
}

knitr::opts_chunk$set(
  echo = FALSE, warning = FALSE, message = FALSE,
  fig.width = 9, fig.height = 5, dpi = 120
)
library(ggplot2)
library(scales)

read_result <- function(file_name) {
  read.csv(
    file.path(project_root, "analysis", "results", file_name),
    stringsAsFactors = FALSE,
    check.names = FALSE
  )
}
monthly <- read_result("washington_monthly_analysis.csv")
monthly$month <- as.Date(monthly$month)
events <- read_result("washington_event_summary.csv")
its <- read_result("washington_its_coefficients.csv")
county <- read_result("washington_county_postwar.csv")
models <- read_result("washington_model_comparison.csv")
gas_models <- read_result("washington_gas_model_comparison.csv")
gas_coefficients <- read_result("washington_gas_model_coefficients.csv")
gas_sensitivity <- read_result("washington_gas_model_sensitivity.csv")
conjoint <- read_result("conjoint_partworths.csv")
conjoint_tests <- read_result("conjoint_attribute_tests.csv")
conjoint_pairwise <- read_result("conjoint_pairwise_contrasts.csv")
conjoint_audit <- read_result("conjoint_data_audit.csv")
data_audit <- read_result("washington_data_audit.csv")
vmt <- read.csv(
  file.path(project_root, "data", "washington_vmt_monthly.csv"),
  stringsAsFactors = FALSE
)
vmt$month <- as.Date(vmt$month)

postwar <- events[events$period == "Post-war", ]
prewar <- events[
  events$period == "Post-incentive-rolloff / pre-war",
]
war_effect <- its[its$term == "post_war", ]
rolloff_effect <- its[
  its$term == "post_combined_incentive_rolloff",
]
gas_effect <- gas_coefficients[
  gas_coefficients$term == "gas_price_prior_3m",
]

paper <- "#f6f1e7"; ink <- "#27241f"; teal <- "#176b68"
rust <- "#b9472e"; gold <- "#8b6b2f"
report_theme <- theme_minimal(base_size = 12) +
  theme(
    plot.background = element_rect(fill = paper, color = NA),
    panel.background = element_rect(fill = paper, color = NA),
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    axis.text = element_text(color = ink),
    axis.title = element_text(color = ink, face = "bold"),
    plot.title = element_text(face = "bold", color = ink),
    legend.position = "top",
    legend.title = element_blank()
  )
```

# Executive report

## Decision question

Washington vehicle manufacturers, dealers, and policy analysts need to know
whether the early-2026 gasoline-price shock produced a meaningful ZEV-demand
rebound after Washington and federal incentives rolled off in 2025. We ask:
**Did Washington's share of new light-duty ZEV title transactions rise after
the war, and was that movement distinguishable from the post-incentive-rolloff
market?**

The answer is cautious. Washington's pooled ZEV-title share increased from
13.9% in November 2025–February 2026 to 14.5% in March–June 2026, a 0.56-point
increase, while average regular gasoline prices rose 33.8%. An interrupted
time-series model estimates an additional post-war change of 1.15 percentage
points, but it is not statistically distinct from zero (Newey–West p = 0.285).
The evidence does not yet show that fuel pressure restored the demand lost
after incentives rolled off. Lagged gasoline prices add modest predictive
information, but the result is sensitive to the electricity-price control.

## Data and definitions

The primary source is the Washington Department of Licensing's public
[Vehicle Title Transactions](https://data.wa.gov/Transportation/Vehicle-Title-Transactions-by-Department-of-Licens/cdk6-5kdf)
API. The pipeline uses 114 consecutive months from January 2017 through June
2026. It requires original titles, the DOL new-vehicle flag, Washington owner
state, passenger cars/multipurpose vehicles/light trucks, and light-duty weight
classes. ZEVs include BEVs, PHEVs, and hydrogen fuel-cell vehicles.

This measure is a registration transaction, not a dealer sale date. Title
processing can lag purchase timing, so October 2025 remains in the
incentive-transition/processing period and the modeled post-rolloff period
begins in November. Washington's clean-vehicle sales-tax exemption ended July
31, 2025, shortly before the federal credit ended September 30, so their
effects cannot be separated reliably. Washington gasoline prices come from the
[U.S. Energy Information Administration](https://www.eia.gov/dnav/pet/hist/LeafHandler.ashx?f=W&n=PET&s=EMM_EPMR_PTE_SWA_DPG).
California Energy Commission quarterly sales are retained only as a neighboring
comparison because their timing and construction differ.

### Data sources, units, and sample sizes

| Dataset | Source | Analysis unit | Coverage | Sample size |
|---|---|---|---|---|
| Washington title market | Washington Department of Licensing | Monthly statewide ZEV-title share | January 2017–June 2026 | 114 months from 2,541,732 titles, including 270,363 ZEV titles |
| Washington county title panel | Washington Department of Licensing | County-month ZEV-title share | January 2017–June 2026 | 4,558 county-month rows and 39 named counties |
| Gas and economic controls | EIA and BLS | Monthly statewide series | January 2017–July 2026 | 115 gas-price, 114 unemployment, and 113 electricity-price rows |
| Washington driving | Federal Highway Administration | Monthly statewide vehicle miles traveled | January 2023–May 2026 | 41 months |
| California comparison | California Energy Commission | Quarterly inferred ZEV-sales share | 2023 Q1–2026 Q2 | 14 quarters |
| Class conjoint survey | Anonymous MSBA 580 class survey | Respondent-profile rating | Final class survey | 18 respondents, 123 usable ratings, and 17 unique profiles |

The regression sample size is 114 monthly observations, not 2.54 million
independent observations. The larger number describes the title transactions
aggregated into those monthly shares. The Shiny app provides a downloadable ZIP
containing the analysis-ready CSV inputs and data dictionary.

## Analytical methods

The first technique is a pre-event regression benchmark trained through June
2025. We compare seasonal-only, linear-time, and quadratic-time models, each
including calendar-month effects and a COVID disruption indicator. Rolling
one-month-ahead validation covers 42 months. The quadratic specification has
the lowest RMSE, 3.66 percentage points, narrowly outperforming the linear
model's 3.78 points.

The second specification is an interrupted time series using all 114 months.
It adds combined post-incentive-rolloff and post-war level indicators to the
quadratic trend, seasonality, and COVID control. Newey–West standard errors
address residual autocorrelation. Event coefficients describe deviations
associated with each period; they do not identify causal effects.

The third specification tests gasoline-price association. Current and lagged
gasoline measures are compared using the same rolling window. The
prior-three-month average performs best and is then fit with trend, seasonality,
COVID, Washington incentive timing, and the combined rolloff period.
Washington unemployment, electricity prices, and a post-war indicator are
added separately as sensitivity checks.

Monthly Washington titles materially improve the study design over the earlier
California-only quarterly approach. They provide enough pre-event history to
estimate recurring seasonal patterns, reserve 42 months for rolling forecast
validation, and distinguish the credit transition from the later war period.
The county panel then tests whether the statewide change was geographically
broad or driven by one large market. California is not pooled into the
regression: it serves only as a directional comparison because California
reports inferred sales while Washington reports title transactions.

### Method strengths and limitations

| Method | Role in this project | Strength | Limitation |
|---|---|---|---|
| Descriptive comparisons | Summarize title share, VMT, county, and California patterns | Transparent measures show observed magnitude and timing | No controls for competing explanations and no causal identification |
| Multiple regression benchmark | Estimate the pre-rolloff path using trend, seasonality, and COVID | Candidate models are compared with 42 rolling one-month forecasts | Historical structure may not remain informative after a market shock |
| Interrupted time series | Estimate adjusted level differences after the incentive rolloff and war | Uses the full monthly history and Newey–West uncertainty | Only four post-war months and overlapping events prevent causal attribution |
| Gas-price regression | Test whether prior-three-month gasoline prices add predictive signal | Lag comparisons, rolling validation, and sensitivity specifications test robustness | Observational estimate weakens after adding electricity prices |
| Rating-based conjoint analysis | Estimate stated preference for price, brand, and fuel economy | Respondent effects isolate within-person tradeoffs across profiles | Only 18 respondents; brand and fuel economy are partially confounded |

Multiple regression and conjoint analysis are the two primary course methods.
The interrupted-time-series and gas-price specifications are extensions of the
multiple-regression method rather than separate course techniques.

```{r benchmark, fig.cap="Observed Washington ZEV-title share and the pre-rolloff benchmark."}
display <- monthly[monthly$month >= as.Date("2021-01-01"), ]
ggplot(display, aes(month)) +
  geom_ribbon(
    aes(ymin = empirical_prediction_low, ymax = empirical_prediction_high),
    fill = teal, alpha = .12
  ) +
  geom_line(aes(y = expected_zev_share, color = "Expected"), linewidth = .9) +
  geom_line(aes(y = zev_share, color = "Observed"), linewidth = 1.05) +
  geom_vline(xintercept = as.Date("2025-11-01"), color = rust, linetype = "dashed") +
  geom_vline(xintercept = as.Date("2026-03-01"), color = rust, linetype = "dotted") +
  scale_color_manual(values = c("Expected" = gold, "Observed" = teal)) +
  scale_y_continuous(labels = label_percent(accuracy = 1)) +
  scale_x_date(date_breaks = "6 months", date_labels = "%Y\n%b") +
  labs(
    title = "ZEV title share remained below its pre-rolloff trajectory",
    x = NULL, y = "New light-duty ZEV title share"
  ) +
  report_theme
```

## Findings

The benchmark expected a pooled 25.7% ZEV-title share during March–June 2026,
11.2 points above the observed 14.5%. The interrupted model estimates a
combined post-incentive-rolloff level shift of −11.59 points (p < 0.001). Its
additional post-war coefficient is positive but uncertain: +1.15 points with a
95% interval from −0.97 to +3.28 points.

Gasoline prices averaged \$3.90 per gallon in the four post-rolloff/pre-war
months and \$5.22 during the four post-war months. Thus, higher fuel costs
coincided with only a small title-share increase. This is inconsistent with a
large immediate fuel-price-driven recovery, although purchasing and title
processing may respond with longer lags.

The prior-three-month gas measure reduces rolling RMSE from 3.66 to 3.37
percentage points, a 7.9% improvement. The policy-adjusted estimate is +2.77
ZEV-share percentage points per \$1/gallon (95% CI +0.27 to +5.27; p = 0.031).
After adding Washington electricity prices, the estimate is +2.00 points and
its interval includes zero (p = 0.118). The result is therefore a modest,
specification-sensitive predictive association, not a causal effect.

FHWA VMT was 1.0% below the same month one year earlier in preliminary May 2026
data. VMT is treated as a separate behavioral outcome rather than a ZEV-share
control because driving can itself respond to gasoline prices.

County patterns were more favorable but heterogeneous. Thirty-two of 39
counties increased ZEV-title share, and the median change was +3.34 points.
Among large markets, Pierce gained 4.70 points, Snohomish 4.51, and Clark 6.90,
while King declined 3.18. County results exclude the DOL's “Unknown or Out of
State” category and should not be treated as independent statewide experiments.

```{r county, fig.cap="Change in ZEV-title share for the 15 largest county markets."}
county_display <- head(county[order(-county$postwar_new_ldv_titles), ], 15)
county_display$county <- reorder(
  county_display$county,
  county_display$change_percentage_points
)
ggplot(county_display, aes(change_percentage_points, county)) +
  geom_vline(xintercept = 0, color = "#746e64") +
  geom_col(aes(fill = change_percentage_points >= 0), show.legend = FALSE) +
  scale_fill_manual(values = c("TRUE" = teal, "FALSE" = rust)) +
  labs(
    title = "Most large counties improved, but King County did not",
    x = "Post-war change (percentage points)", y = NULL
  ) +
  report_theme
```

The rating-based conjoint survey adds stated-preference evidence from 123
ratings supplied by 18 respondents. Price is the clearest attribute
(cluster-robust p = 0.013): a \$30,000 profile scores 1.30 preference points
above a \$120,000 profile (p = 0.005). Brand and fuel economy cannot be cleanly
separated because Tesla was consistently paired with 110 MPGe.

## Recommendations and limitations

Managers should avoid a broad Washington inventory expansion based on four
post-war months. Prioritize affordability and total-cost messaging, then monitor
whether title share continues improving as gasoline prices remain elevated.
County-level campaigns are more defensible in markets showing both scale and
positive movement, such as Pierce, Snohomish, and Clark.

The study is observational. Washington and federal policy, manufacturer
incentives, interest rates, vehicle prices, model availability, title-processing
delays, and the war overlap. The state and federal incentive expirations are
especially difficult to separate. The post-war period is short, the
gas-association estimate weakens with electricity prices, rolling forecast
error is material, and the conjoint sample is small and partially confounded.
Later DOL releases are needed before claiming a durable rebound.

## Sources

- [Washington Department of Licensing title transactions](https://data.wa.gov/Transportation/Vehicle-Title-Transactions-by-Department-of-Licens/cdk6-5kdf)
- [EIA Washington regular gasoline prices](https://www.eia.gov/dnav/pet/hist/LeafHandler.ashx?f=W&n=PET&s=EMM_EPMR_PTE_SWA_DPG)
- [BLS Washington unemployment](https://data.bls.gov/timeseries/LAUST530000000000003)
- [FHWA Traffic Volume Trends](https://www.fhwa.dot.gov/policyinformation/travel_monitoring/tvt.cfm)
- [Washington clean-vehicle sales-tax exemption](https://dor.wa.gov/forms-publications/publications-subject/special-notices/new-clean-alternative-fuel-and-plug-hybrid-vehicle-sales-and-use-tax-exemption)
- [Washington EV Instant Rebate](https://www.commerce.wa.gov/clean-transportation/ev-instant-rebate/)
- [California Energy Commission ZEV sales](https://www.energy.ca.gov/data-reports/energy-almanac/zero-emission-vehicle-and-infrastructure-statistics-collection/new-zev)

# Technical calculation appendix

This appendix shows how each result is calculated and how each variable enters
the analysis. It is supplemental to the executive report. All outcome shares
are stored as proportions in the model, then multiplied by 100 when reported
as percentage points.

## Calculation map

| Study | Outcome being explained | Main inputs | What the result means |
|---|---|---|---|
| Descriptive event comparison | Pooled Washington ZEV-title share | ZEV titles, all new light-duty titles, event window | Observed before-versus-after difference; no controls |
| Pre-rolloff benchmark | Monthly Washington ZEV-title share | Time, time squared, calendar month, COVID period | Expected share if the pre-rolloff pattern continued |
| Interrupted time series | Monthly Washington ZEV-title share | Benchmark variables, combined incentive rolloff, post-event indicator | Adjusted level differences associated with the two periods |
| Gas-price model | Monthly Washington ZEV-title share | Prior-three-month gas price and policy, trend, seasonal, and COVID controls | Predictive association per \$1 per gallon |
| VMT response | Year-over-year Washington vehicle miles traveled | Current-month and prior-year VMT | Separate driving response, not a ZEV-share control |
| County comparison | Pooled county ZEV-title share | County titles in the two four-month windows | Geographic heterogeneity in percentage points |
| California comparison | Indexed quarterly ZEV share | Washington titles and California inferred sales | Directional comparison from a common baseline |
| Conjoint survey | Reversed profile rating | Price, brand, fuel economy, respondent | Stated preference associated with each attribute level |

## 1. Primary ZEV-title share

For month $t$, the primary outcome is:

$$
ZEVShare_t =
\frac{BEV_t + PHEV_t + FCEV_t}
{All\ new\ light\text{-}duty\ original\ titles_t}.
$$

The numerator and denominator use the same filters: original title, DOL `New`
flag, Washington owner state, eligible passenger/light-truck vehicle types,
and light-duty weight classes. This makes powertrain classification the only
difference between the numerator and denominator.

Across the complete study period, the pipeline contains
`r format(as.numeric(data_audit$value[data_audit$metric == "total_new_ldv_titles"]), big.mark = ",")`
new light-duty titles and
`r format(as.numeric(data_audit$value[data_audit$metric == "total_zev_titles"]), big.mark = ",")`
ZEV titles over
`r data_audit$value[data_audit$metric == "consecutive_months"]` consecutive
months.

## 2. Descriptive event-window calculation

The pooled share gives months with more title transactions proportionally more
weight:

$$
PooledShare_w = \frac{\sum_{t \in w} ZEVTitles_t}
{\sum_{t \in w} NewLDVTitles_t}.
$$

```{r event-window-calculation}
event_calculation <- data.frame(
  Window = c("November 2025-February 2026", "March-June 2026"),
  `ZEV titles` = c(prewar$zev_titles, postwar$zev_titles),
  `All new LDV titles` = c(prewar$new_ldv_titles, postwar$new_ldv_titles),
  Calculation = c(
    paste0(prewar$zev_titles, " / ", prewar$new_ldv_titles),
    paste0(postwar$zev_titles, " / ", postwar$new_ldv_titles)
  ),
  `Pooled share` = percent(
    c(prewar$pooled_zev_share, postwar$pooled_zev_share),
    accuracy = 0.1
  ),
  check.names = FALSE
)
knitr::kable(event_calculation, caption = "Observed pooled-share calculation")
```

The observed change is:

$$
(0.14479 - 0.13922) \times 100 = 0.56
\text{ percentage points}.
$$

Gasoline prices changed from \$3.90 to \$5.22 per gallon:

$$
\left(\frac{5.2213}{3.9019} - 1\right) \times 100 = 33.8\%.
$$

These are descriptive calculations. They do not adjust for trend, seasonality,
policy, or other concurrent changes.

## 3. Pre-rolloff regression benchmark

The selected benchmark is estimated only through June 2025:

$$
ZEVShare_t = \beta_0 + \beta_1 Time_t + \beta_2 Time_t^2
+ \sum_{m=2}^{12}\gamma_m Month_{m,t}
+ \delta COVID_t + \epsilon_t.
$$

| Variable | How it enters | What it changes |
|---|---|---|
| `Time` | Consecutive month number | Long-run adoption trajectory |
| `Time squared` | Squared month number | Allows the trajectory to accelerate or flatten |
| Calendar month indicators | February through December, relative to January | Recurring seasonality |
| `COVID` | 1 from March 2020 through June 2021 | Broad pandemic-era level difference |

The model compares its one-month-ahead prediction $\hat{y}_{t|t-1}$ with the
observed share in each of 42 rolling validation months:

$$
RMSE = \sqrt{\frac{1}{42}\sum_{t=1}^{42}
(y_t - \hat{y}_{t|t-1})^2}.
$$

```{r benchmark-model-comparison}
benchmark_table <- models[, c("model", "parameters", "rolling_rmse", "rolling_mae", "selected")]
benchmark_table$rolling_rmse <- percent(benchmark_table$rolling_rmse, accuracy = 0.01)
benchmark_table$rolling_mae <- percent(benchmark_table$rolling_mae, accuracy = 0.01)
names(benchmark_table) <- c("Model", "Parameters", "Rolling RMSE", "Rolling MAE", "Selected")
knitr::kable(benchmark_table, caption = "Time-ordered benchmark comparison")
```

The quadratic model has the lowest rolling RMSE, 3.66 percentage points. Its
March-June 2026 fitted values average 25.7%, compared with 14.5% observed:

$$
(0.14479 - 0.25719) \times 100 = -11.24
\text{ percentage points}.
$$

The dashboard band is $\hat{y}_t \pm 1.96 \times RMSE$, or approximately
$\hat{y}_t \pm 7.2$ percentage points. It is an empirical forecast-error band,
not a causal confidence interval.

## 4. Interrupted time-series calculation

The event model adds two level indicators to the benchmark:

$$
ZEVShare_t = BenchmarkTerms_t
+ \tau_R Rolloff_t + \tau_E PostEvent_t + \epsilon_t.
$$

`Rolloff` equals 1 beginning November 2025. `PostEvent` equals 1 beginning
March 2026, so $\tau_E$ is the additional level difference after March,
conditional on the combined rolloff, trend, seasonality, and COVID terms.

```{r event-coefficient-calculation}
event_terms <- its[its$term %in% c(
  "post_combined_incentive_rolloff", "post_war"
), c("term", "estimate", "confidence_low", "confidence_high", "p_value")]
event_terms$estimate <- round(100 * event_terms$estimate, 2)
event_terms$confidence_low <- round(100 * event_terms$confidence_low, 2)
event_terms$confidence_high <- round(100 * event_terms$confidence_high, 2)
event_terms$p_value <- format.pval(event_terms$p_value, digits = 3, eps = 0.001)
names(event_terms) <- c("Variable", "Effect (percentage points)", "CI low", "CI high", "p-value")
knitr::kable(event_terms, caption = "How the event indicators affect modeled ZEV-title share")
```

For example, the post-event coefficient is converted to percentage points as:

$$
0.011523 \times 100 = 1.15 \text{ percentage points}.
$$

Newey-West standard errors with four lags adjust the confidence intervals and
p-values for monthly heteroskedasticity and serial correlation. They do not
change the coefficient estimates. Because the indicators are not randomized,
the coefficients are adjusted associations, not causal effects.

## 5. Gas-price association calculation

The exposure precedes the title month:

$$
Prior3MonthGas_t = \frac{Gas_{t-1} + Gas_{t-2} + Gas_{t-3}}{3}.
$$

The association model is:

$$
ZEVShare_t = BenchmarkTerms_t + PolicyTerms_t
+ \theta Prior3MonthGas_t + \epsilon_t.
$$

The policy terms indicate Washington's sales-tax exemption, the instant-rebate
period, the July-October 2025 transition, and the combined post-rolloff period.
Unemployment, electricity price, and the post-event indicator are added in
separate sensitivity specifications.

```{r gas-sensitivity-calculation}
gas_display <- gas_sensitivity[, c(
  "model", "gas_estimate", "gas_confidence_low", "gas_confidence_high", "gas_p_value"
)]
gas_display$gas_estimate <- round(100 * gas_display$gas_estimate, 2)
gas_display$gas_confidence_low <- round(100 * gas_display$gas_confidence_low, 2)
gas_display$gas_confidence_high <- round(100 * gas_display$gas_confidence_high, 2)
gas_display$gas_p_value <- format.pval(gas_display$gas_p_value, digits = 3, eps = 0.001)
names(gas_display) <- c("Specification", "Effect per USD 1 (points)", "CI low", "CI high", "p-value")
knitr::kable(gas_display, caption = "Gas-price coefficient across control specifications")
```

The policy-adjusted coefficient is calculated as
$0.027701 \times 100 = 2.77$ percentage points per \$1 per gallon. Adding
electricity prices reduces it to 2.00 points and widens the uncertainty across
zero. This sensitivity is why the result is described as predictive rather
than causal.

## 6. Vehicle-miles-traveled calculation

VMT is evaluated as a separate outcome:

$$
VMTYoY_t = \left(\frac{VMT_t}{VMT_{t-12}} - 1\right) \times 100.
$$

A value of -1.0% means estimated travel was 1.0% lower than in the same month
one year earlier. VMT is not included as a ZEV-share predictor because driving
may itself respond to gasoline prices.

## 7. County change calculation

For each county $c$:

$$
CountyChange_c = 100 \times
\left(
\frac{PostEventZEVTitles_c}{PostEventLDVTitles_c}
-
\frac{PreEventZEVTitles_c}{PreEventLDVTitles_c}
\right).
$$

The median county change gives every county equal weight. The statewide change
weights counties by title volume. This explains how 32 of 39 counties can
increase while the statewide gain remains 0.56 points, especially because
large King County declined.

## 8. California index calculation

Washington titles and California inferred sales have different definitions,
so the dashboard compares direction after setting each series to 100 in 2025
Q4:

$$
Index_{s,q} = \frac{ZEVShare_{s,q}}{ZEVShare_{s,2025Q4}} \times 100.
$$

An index of 105 means the share is 5% above that state's own 2025 Q4 value. It
does not mean the two states have the same ZEV-share level.

## 9. Conjoint calculation

The original survey scale uses 1 as best and 5 as worst. It is reversed so
higher values mean stronger preference:

$$
PreferenceScore_{i,j} = 6 - Rating_{i,j}.
$$

The rating model is:

$$
PreferenceScore_{i,j} = \alpha_i
+ Brand_j + FuelEconomy_j + Price_j + \epsilon_{i,j},
$$

where $\alpha_i$ is a respondent fixed effect. Brand, fuel economy, and price
are effect-coded, so each level is interpreted relative to its attribute's
grand mean. CR1 standard errors are clustered by respondent because each
person rates multiple profiles.

```{r conjoint-calculation}
price_test <- conjoint_tests[conjoint_tests$attribute == "Price", ]
price_contrast <- conjoint_pairwise[
  conjoint_pairwise$attribute == "Price" &
    conjoint_pairwise$level_1 == "$30,000" &
    conjoint_pairwise$level_2 == "$120,000",
]
conjoint_calculation <- data.frame(
  Result = c(
    "Rated respondents",
    "Usable profile ratings",
    "Joint price test",
    "USD 30,000 minus USD 120,000"
  ),
  Calculation = c(
    conjoint_audit$value[conjoint_audit$metric == "survey_ids_with_at_least_one_rating"],
    conjoint_audit$value[conjoint_audit$metric == "complete_ratings"],
    paste0("F = ", round(price_test$f_statistic, 2)),
    paste0(round(price_contrast$preference_difference, 2), " preference points")
  ),
  `p-value` = c(
    NA,
    NA,
    round(price_test$p_value, 3),
    round(price_contrast$p_value, 3)
  ),
  check.names = FALSE
)
knitr::kable(conjoint_calculation, caption = "Conjoint sample and price calculation")
```

Price is the clearest survey attribute. Brand and fuel economy are not given
standalone managerial interpretations because Tesla is not crossed with the
lower fuel-economy levels, which confounds those two variables.

## Variable-to-decision summary

| Variable movement | Supported interpretation | Decision implication |
|---|---|---|
| Post-event indicator is positive but uncertain | No distinct recovery is established | Do not expand statewide inventory on this result alone |
| Higher prior-three-month gas price predicts higher ZEV share | Fuel cost adds modest predictive information | Use total-cost messaging, but do not claim causation |
| Electricity control weakens the gas coefficient | Relative energy costs and correlated conditions matter | Treat the gas result as specification-sensitive |
| Most counties increase but King declines | Statewide averages mask geography | Target large positive markets selectively |
| Lower survey price raises preference | Affordability is the clearest stated preference | Lead with price and affordability |
| VMT is mixed | No sustained driving reduction is established | Avoid overstating behavioral response |
