---
title: "California EV Demand After Two Market Shocks"
subtitle: "Tax-credit expiration, fuel pressure, and the 2026 rebound"
author: "MSBA 580 Project Group"
date: "`r format(Sys.Date(), '%B %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    self_contained: true
---

```{r setup, include=FALSE}
report_path <- normalizePath(knitr::current_input(dir = TRUE), mustWork = TRUE)
project_root <- normalizePath(file.path(dirname(report_path), ".."), mustWork = TRUE)
analysis_script <- file.path(project_root, "scripts", "run_descriptive_analysis.R")
analysis_status <- system2(file.path(R.home("bin"), "Rscript"), analysis_script)
if (analysis_status != 0L) stop("The analysis pipeline did not complete.")

knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.width = 9,
  fig.height = 5,
  dpi = 120
)
library(ggplot2)
library(scales)

read_result <- function(file_name) {
  read.csv(
    file.path(project_root, "analysis", "results", file_name),
    stringsAsFactors = FALSE,
    check.names = FALSE
  )
}

quarterly <- read.csv(
  file.path(project_root, "data", "quarterly_controls.csv"),
  stringsAsFactors = FALSE
)
quarterly$quarter_start <- as.Date(quarterly$quarter_start)
benchmark <- read_result("seasonal_benchmark.csv")
benchmark$quarter_start <- as.Date(benchmark$quarter_start)
segments <- read_result("county_segments.csv")
segment_profiles <- read_result("county_segment_profiles.csv")
model_comparison <- read_result("regression_model_comparison.csv")
diagnostics <- read_result("regression_diagnostics.csv")
vmt <- read.csv(
  file.path(project_root, "data", "california_vmt_monthly.csv"),
  stringsAsFactors = FALSE
)
vmt$observation_month <- as.Date(vmt$observation_month)

paper <- "#f6f1e7"
ink <- "#27241f"
teal <- "#176b68"
rust <- "#b9472e"
gold <- "#8b6b2f"

report_theme <- theme_minimal(base_size = 12) +
  theme(
    plot.background = element_rect(fill = paper, color = NA),
    panel.background = element_rect(fill = paper, color = NA),
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    axis.text = element_text(color = ink),
    axis.title = element_text(color = ink, face = "bold"),
    plot.title = element_text(face = "bold", color = ink),
    legend.position = "top",
    legend.title = element_blank()
  )

quarter_date_label <- function(x) {
  paste0(format(x, "%Y"), " Q", ((as.integer(format(x, "%m")) - 1L) %/% 3L) + 1L)
}
```

# Executive report

## Decision question

California EV manufacturers, dealers, and marketing managers need to decide
whether the early-2026 gasoline-price shock created a durable EV sales
opportunity after the federal purchase credit expired in September 2025. We ask:
**Did California's zero-emission vehicle (ZEV) market rebound as fuel pressure
increased, and was the rebound large enough to offset the post-credit decline?**
We also examine whether Californians reduced driving and whether counties
followed distinct response patterns.

The answer is cautious. California's official ZEV share recovered from its
2026 Q1 low, but by 2026 Q2 it was only 0.19 percentage points above the
immediate post-credit 2025 Q4 level. The rebound coincided with sharply higher
gasoline prices and much greater EV search interest, yet the quarterly record
cannot show that either caused purchases. Vehicle-miles traveled (VMT) provides
mixed evidence of reduced driving. Managers should treat the period as renewed
consumer attention, not proof of a permanent demand shift.

## Data and analytical approach

The project combines independently sourced public data. California Energy
Commission records provide quarterly new light-duty ZEV and total light-duty
sales statewide and by county. ZEVs include battery electric, plug-in hybrid,
and hydrogen fuel-cell vehicles; they exclude used and medium- or heavy-duty
vehicles. U.S. Energy Information Administration data supply California regular
gasoline prices. Google Trends supplies a normalized relative search-interest
index, not search counts. FHWA Traffic Volume Trends, based on state-reported
counts, supplies monthly statewide VMT. Economic controls are retained in the
project dataset for context.

We use two major analytical techniques. First, a pre-event regression benchmark
fits a linear time trend and calendar-quarter indicators to 2023 Q1 through
2025 Q2, then compares later observations with predictions. We compare that
specification with mean-only, time-only, and seasonal-only alternatives using
small-sample AICc and leave-one-out error. We also report HC1 robust standard
errors, residual normality, an approximate Durbin-Watson statistic, Cook's
distance, and leave-one-out RMSE.

Second, k-means segmentation groups the 58 California county mailing-address
categories using standardized 2025 Q4 ZEV share, the change through 2026 Q2,
and log average quarterly light-duty sales. We inspect two through six clusters
using mean silhouette width and select three for parsimony and business
interpretability. The silhouette near 0.34 indicates only moderate separation,
so these are descriptive archetypes rather than permanent county types.

## What changed?

ZEV share reached 28.3% in 2025 Q3 as purchases were likely pulled forward
before the federal credit expired. It then fell 9.43 percentage points to 18.9%
in 2025 Q4, a 33.3% relative decline, and fell again to 15.8% in 2026 Q1. The
share rebounded 3.33 percentage points to 19.1% in 2026 Q2. Thus, the full
post-war quarter recovered the first-quarter loss but did not meaningfully
exceed the immediate post-credit baseline.

```{r statewide-benchmark, fig.cap="Observed statewide ZEV share and the pre-event time-plus-season benchmark."}
ggplot(benchmark, aes(quarter_start)) +
  geom_ribbon(
    aes(ymin = prediction_low, ymax = prediction_high),
    fill = teal, alpha = 0.12
  ) +
  geom_line(aes(y = expected_zev_share), color = gold, linewidth = 0.9) +
  geom_line(aes(y = observed_zev_share), color = teal, linewidth = 1.1) +
  geom_point(aes(y = observed_zev_share), color = rust, size = 2.4) +
  geom_vline(
    xintercept = as.Date("2025-09-30"),
    color = rust, linetype = "dashed"
  ) +
  scale_y_continuous(labels = label_percent(accuracy = 1)) +
  scale_x_date(date_breaks = "6 months", labels = quarter_date_label) +
  labs(
    title = "The 2026 Q2 rebound remained below the benchmark",
    subtitle = "Shading is a 95% prediction interval; dashed line marks credit expiration",
    x = NULL,
    y = "ZEV share"
  ) +
  report_theme
```

The benchmark expected a 22.5% share in 2026 Q2, 3.38 percentage points above
the observed value, although the observation remained inside a wide 95%
prediction interval of 17.7% to 27.3%. Diagnostics reinforce caution. The model
has only 10 training quarters and five residual degrees of freedom. One
observation exceeds the common Cook's-distance screening threshold. Moreover,
the simpler seasonal-only model has lower leave-one-out RMSE, while the
mean-only model has the best AICc. The benchmark is useful for visualization,
not a credible causal counterfactual.

Between 2025 Q4 and 2026 Q2, average California regular gasoline prices rose
32.0%, from $4.34 to $5.73 per gallon, and Google Trends interest in “electric
vehicle” rose 140%. Attention clearly increased, but a normalized search index
cannot measure the number or purchase intent of searchers.

## Geographic and behavioral response

Forty-nine of 58 counties increased ZEV share, with a median increase of 2.35
percentage points. The official statewide result includes an “Out of State”
mailing-address category; excluding that category raises the aggregate change
from 0.19 to 1.27 percentage points. The official measure remains primary, and
the county-only calculation is a sensitivity check.

```{r county-segments, fig.cap="County segment positions; point size represents average quarterly light-duty sales."}
ggplot(
  segments,
  aes(
    100 * zev_share_2025_q4,
    change_percentage_points,
    size = average_quarterly_ldv_sales,
    color = segment
  )
) +
  geom_hline(yintercept = 0, color = "#746e64", linetype = "dotted") +
  geom_point(alpha = 0.8) +
  scale_size_area(max_size = 13, labels = label_comma()) +
  scale_color_manual(values = c(
    "Large established EV markets" = teal,
    "Low-adoption / stalled" = rust,
    "Small-market / fast rebound" = gold
  )) +
  labs(
    title = "County markets followed three broad response patterns",
    x = "2025 Q4 ZEV share (%)",
    y = "Change to 2026 Q2 (percentage points)",
    size = "Average quarterly sales"
  ) +
  report_theme
```

The large established segment contains 27 counties, averages a 19.8% baseline
share, and gains 2.03 percentage points. The low-adoption/stalled segment
contains 16 counties and declines 0.36 points on average. Fifteen small-market
counties form a fast-rebound segment averaging a 6.09-point increase, but their
small sales denominators make large movements more likely.

VMT does not show a large sustained driving reduction. March, April, and May
2026 changed +3.5%, -0.8%, and -0.1% from the same months a year earlier. The
three-month mean was +0.87%; May is preliminary and June was unavailable.

```{r vmt-response, fig.cap="Monthly change from the same month one year earlier."}
vmt_display <- vmt[vmt$observation_month >= as.Date("2024-01-01"), ]
vmt_display$direction <- ifelse(
  vmt_display$yoy_change >= 0, "More driving", "Less driving"
)
ggplot(vmt_display, aes(observation_month, yoy_change, fill = direction)) +
  geom_hline(yintercept = 0, color = "#746e64") +
  geom_col(width = 24) +
  geom_vline(
    xintercept = as.Date("2026-02-28"),
    color = rust, linetype = "dashed"
  ) +
  scale_fill_manual(values = c("More driving" = teal, "Less driving" = rust)) +
  scale_y_continuous(labels = label_percent(accuracy = 1)) +
  scale_x_date(date_breaks = "3 months", date_labels = "%Y\n%b") +
  labs(
    title = "Early post-war driving evidence is mixed",
    x = NULL,
    y = "Year-over-year VMT change"
  ) +
  report_theme
```

## Managerial recommendations

Managers should not make a statewide inventory bet based on one post-war
quarter. In large established EV markets, maintain inventory availability and
use fuel-cost messaging to convert renewed search interest. In low-adoption
counties, diagnose charging access, affordability, and awareness before adding
inventory. In small fast-rebound markets, use light-touch campaigns and avoid
overreacting to volatile percentage changes. Continue monthly leading-indicator
tracking and re-estimate the benchmark when more quarters become available.

## Limitations and next analysis

The study is observational and cannot separate the war, gasoline prices, tax
policy, interest rates, manufacturer incentives, seasonality, and vehicle
prices. County observations share the same statewide environment and are not
independent replications of the shock. Google Trends is relative, VMT is
incomplete for 2026 Q2, and the sales series has only one complete post-war
quarter.

**Conjoint analysis is intentionally pending.** When the class survey data are
available, the project will estimate attribute utilities, relative importance,
willingness to pay, and scenario choice shares. No conjoint findings are
included in the current conclusions.

## Sources

- [California Energy Commission ZEV sales](https://www.energy.ca.gov/data-reports/energy-almanac/zero-emission-vehicle-and-infrastructure-statistics-collection/new-zev)
- [U.S. Energy Information Administration gasoline prices](https://www.eia.gov/dnav/pet/pet_pri_gnd_dcus_sca_w.htm)
- [Google Trends](https://trends.google.com/trends/)
- [FHWA Traffic Volume Trends](https://www.fhwa.dot.gov/policyinformation/travel_monitoring/tvt.cfm)
