---
title: "California EV Demand After Two Market Shocks"
subtitle: "Tax-credit expiration, fuel pressure, and the 2026 rebound"
author: "MSBA 580 Project Group"
date: "`r format(Sys.Date(), '%B %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    self_contained: true
---

```{r setup, include=FALSE}
report_path <- normalizePath(knitr::current_input(dir = TRUE), mustWork = TRUE)
project_root <- normalizePath(file.path(dirname(report_path), ".."), mustWork = TRUE)
analysis_script <- file.path(project_root, "scripts", "run_descriptive_analysis.R")
analysis_status <- system2(file.path(R.home("bin"), "Rscript"), analysis_script)
if (analysis_status != 0L) stop("The analysis pipeline did not complete.")
conjoint_script <- file.path(project_root, "scripts", "run_conjoint_analysis.R")
conjoint_status <- system2(file.path(R.home("bin"), "Rscript"), conjoint_script)
if (conjoint_status != 0L) stop("The conjoint pipeline did not complete.")

knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.width = 9,
  fig.height = 5,
  dpi = 120
)
library(ggplot2)
library(scales)

read_result <- function(file_name) {
  read.csv(
    file.path(project_root, "analysis", "results", file_name),
    stringsAsFactors = FALSE,
    check.names = FALSE
  )
}

quarterly <- read.csv(
  file.path(project_root, "data", "quarterly_controls.csv"),
  stringsAsFactors = FALSE
)
quarterly$quarter_start <- as.Date(quarterly$quarter_start)
benchmark <- read_result("seasonal_benchmark.csv")
benchmark$quarter_start <- as.Date(benchmark$quarter_start)
segments <- read_result("county_segments.csv")
segment_profiles <- read_result("county_segment_profiles.csv")
model_comparison <- read_result("regression_model_comparison.csv")
diagnostics <- read_result("regression_diagnostics.csv")
conjoint_partworths <- read_result("conjoint_partworths.csv")
conjoint_tests <- read_result("conjoint_attribute_tests.csv")
conjoint_pairwise <- read_result("conjoint_pairwise_contrasts.csv")
vmt <- read.csv(
  file.path(project_root, "data", "california_vmt_monthly.csv"),
  stringsAsFactors = FALSE
)
vmt$observation_month <- as.Date(vmt$observation_month)

paper <- "#f6f1e7"
ink <- "#27241f"
teal <- "#176b68"
rust <- "#b9472e"
gold <- "#8b6b2f"

report_theme <- theme_minimal(base_size = 12) +
  theme(
    plot.background = element_rect(fill = paper, color = NA),
    panel.background = element_rect(fill = paper, color = NA),
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    axis.text = element_text(color = ink),
    axis.title = element_text(color = ink, face = "bold"),
    plot.title = element_text(face = "bold", color = ink),
    legend.position = "top",
    legend.title = element_blank()
  )

quarter_date_label <- function(x) {
  paste0(format(x, "%Y"), " Q", ((as.integer(format(x, "%m")) - 1L) %/% 3L) + 1L)
}
```

# Executive report

## Decision question

California EV manufacturers, dealers, and marketing managers need to decide
whether the early-2026 gasoline-price shock created a durable EV sales
opportunity after the federal purchase credit expired in September 2025. We ask:
**Did California's zero-emission vehicle (ZEV) market rebound as fuel pressure
increased, and was the rebound large enough to offset the post-credit decline?**
We also examine whether Californians reduced driving and whether counties
followed distinct response patterns.

California's ZEV share recovered from its 2026 Q1 low, but by Q2 it was only
0.19 percentage points above the post-credit 2025 Q4 level. The rebound
coincided with higher gasoline prices and EV search interest, but cannot prove
causation. VMT provides mixed evidence of reduced driving. Managers should
treat the period as renewed attention, not a permanent demand shift.

## Data and analytical approach

The project combines independently sourced public data. California Energy
Commission records provide quarterly new light-duty ZEV and total light-duty
sales statewide and by county. ZEVs include battery electric, plug-in hybrid,
and hydrogen fuel-cell vehicles; they exclude used and medium- or heavy-duty
vehicles. EIA data supply gasoline prices; Google Trends supplies relative
interest, not search counts; and FHWA Traffic Volume Trends supplies monthly
statewide VMT.

The observed-market analysis uses two major techniques. First, a pre-event regression benchmark
fits a linear time trend and calendar-quarter indicators to 2023 Q1 through
2025 Q2, then compares later observations with predictions. We compare that
specification with simpler alternatives using small-sample AICc and
leave-one-out error and report robust and influence diagnostics.

Second, k-means segmentation groups the 58 California county mailing-address
categories using standardized 2025 Q4 ZEV share, the change through 2026 Q2,
and log average quarterly light-duty sales. We inspect two through six clusters
using mean silhouette width and select three for parsimony and business
interpretability. The silhouette near 0.34 indicates only moderate separation,
so these are descriptive archetypes rather than permanent county types.

Third, a rating-based conjoint model analyzes 123 completed profile ratings
from 18 respondents. Ratings are reversed so higher scores mean greater
preference, then modeled with effect-coded attributes, respondent fixed
effects, and respondent-clustered standard errors.

## What changed?

ZEV share reached 28.3% in 2025 Q3 as purchases were likely pulled forward
before the federal credit expired. It then fell 9.43 percentage points to 18.9%
in 2025 Q4, a 33.3% relative decline, and fell again to 15.8% in 2026 Q1. The
share rebounded 3.33 percentage points to 19.1% in 2026 Q2. Thus, the full
post-war quarter recovered the first-quarter loss but did not meaningfully
exceed the immediate post-credit baseline.

```{r statewide-benchmark, fig.cap="Observed statewide ZEV share and the pre-event time-plus-season benchmark."}
ggplot(benchmark, aes(quarter_start)) +
  geom_ribbon(
    aes(ymin = prediction_low, ymax = prediction_high),
    fill = teal, alpha = 0.12
  ) +
  geom_line(aes(y = expected_zev_share), color = gold, linewidth = 0.9) +
  geom_line(aes(y = observed_zev_share), color = teal, linewidth = 1.1) +
  geom_point(aes(y = observed_zev_share), color = rust, size = 2.4) +
  geom_vline(
    xintercept = as.Date("2025-09-30"),
    color = rust, linetype = "dashed"
  ) +
  scale_y_continuous(labels = label_percent(accuracy = 1)) +
  scale_x_date(date_breaks = "6 months", labels = quarter_date_label) +
  labs(
    title = "The 2026 Q2 rebound remained below the benchmark",
    subtitle = "Shading is a 95% prediction interval; dashed line marks credit expiration",
    x = NULL,
    y = "ZEV share"
  ) +
  report_theme
```

The benchmark expected 22.5% in 2026 Q2, 3.38 points above observed, although
the observation remained inside its 17.7%–27.3% prediction interval. With only
10 training quarters, one influential observation, and simpler models
performing better on some criteria, this is a visualization—not a causal
counterfactual.

Between 2025 Q4 and 2026 Q2, average California regular gasoline prices rose
32.0%, from $4.34 to $5.73 per gallon, and Google Trends interest in “electric
vehicle” rose 140%. Attention clearly increased, but a normalized search index
cannot measure the number or purchase intent of searchers.

## Geographic and behavioral response

Forty-nine of 58 counties increased ZEV share, with a median increase of 2.35
percentage points. The official statewide result includes an “Out of State”
mailing-address category; excluding that category raises the aggregate change
from 0.19 to 1.27 percentage points. The official measure remains primary, and
the county-only calculation is a sensitivity check.

```{r county-segments, fig.cap="County segment positions; point size represents average quarterly light-duty sales."}
ggplot(
  segments,
  aes(
    100 * zev_share_2025_q4,
    change_percentage_points,
    size = average_quarterly_ldv_sales,
    color = segment
  )
) +
  geom_hline(yintercept = 0, color = "#746e64", linetype = "dotted") +
  geom_point(alpha = 0.8) +
  scale_size_area(max_size = 13, labels = label_comma()) +
  scale_color_manual(values = c(
    "Large established EV markets" = teal,
    "Low-adoption / stalled" = rust,
    "Small-market / fast rebound" = gold
  )) +
  labs(
    title = "County markets followed three broad response patterns",
    x = "2025 Q4 ZEV share (%)",
    y = "Change to 2026 Q2 (percentage points)",
    size = "Average quarterly sales"
  ) +
  report_theme
```

The 27 large established markets average a 2.03-point gain. The 16
low-adoption/stalled counties decline 0.36 points. Fifteen small markets average
a 6.09-point rebound, but small denominators magnify movements.

VMT does not show a large sustained driving reduction. March, April, and May
2026 changed +3.5%, -0.8%, and -0.1% from the same months a year earlier. The
three-month mean was +0.87%; May is preliminary and June was unavailable.

```{r vmt-response, fig.cap="Monthly change from the same month one year earlier."}
vmt_display <- vmt[vmt$observation_month >= as.Date("2024-01-01"), ]
vmt_display$direction <- ifelse(
  vmt_display$yoy_change >= 0, "More driving", "Less driving"
)
ggplot(vmt_display, aes(observation_month, yoy_change, fill = direction)) +
  geom_hline(yintercept = 0, color = "#746e64") +
  geom_col(width = 24) +
  geom_vline(
    xintercept = as.Date("2026-02-28"),
    color = rust, linetype = "dashed"
  ) +
  scale_fill_manual(values = c("More driving" = teal, "Less driving" = rust)) +
  scale_y_continuous(labels = label_percent(accuracy = 1)) +
  scale_x_date(date_breaks = "3 months", date_labels = "%Y\n%b") +
  labs(
    title = "Early post-war driving evidence is mixed",
    x = NULL,
    y = "Year-over-year VMT change"
  ) +
  report_theme
```

## Stated preferences

The survey provides complementary stated-preference evidence. Price was the
clearest attribute (cluster-robust p = 0.013): a $30,000 profile scored 1.30
points above a $120,000 profile (p = 0.005). Tesla was paired almost exclusively
with 110 MPGe, so brand and fuel economy cannot be cleanly separated.

```{r conjoint-utilities, fig.cap="Rating-based conjoint part-worth utilities with respondent-clustered 95% intervals."}
conjoint_partworths$level <- factor(
  conjoint_partworths$level,
  levels = rev(conjoint_partworths$level)
)
ggplot(conjoint_partworths, aes(utility, level, color = attribute)) +
  geom_vline(xintercept = 0, color = "#746e64", linetype = "dotted") +
  geom_errorbar(
    aes(xmin = confidence_low, xmax = confidence_high),
    width = 0.16,
    orientation = "y"
  ) +
  geom_point(size = 2.7) +
  facet_wrap(~attribute, scales = "free_y", ncol = 1) +
  scale_color_manual(values = c(
    "Brand" = teal, "Fuel economy" = gold, "Price" = rust
  )) +
  labs(
    title = "Affordability produced the clearest preference signal",
    subtitle = "Positive utilities indicate greater stated preference",
    x = "Part-worth utility (preference-score points)",
    y = NULL
  ) +
  report_theme
```

## Managerial recommendations

Managers should not make a statewide inventory bet based on one post-war
quarter. In large established EV markets, maintain inventory availability and
use fuel-cost messaging to convert renewed search interest. In low-adoption
counties, diagnose charging access, affordability, and awareness before adding
inventory. In small fast-rebound markets, use light-touch campaigns and avoid
overreacting to volatile percentage changes. The survey supports emphasizing
affordability and total ownership cost, but not a brand-specific conclusion.
Continue monthly leading-indicator tracking and re-estimate the benchmark when
more quarters become available.

## Limitations and next analysis

The study is observational and cannot separate the war, gasoline prices, tax
policy, interest rates, manufacturer incentives, seasonality, and vehicle
prices. County observations share the same statewide environment and are not
independent replications of the shock. Google Trends is relative, VMT is
incomplete for 2026 Q2, and the sales series has only one complete post-war
quarter. The conjoint sample is small, uses ratings rather than forced choices,
and has 17 missing ratings. Its brand–fuel-economy confounding prevents clean
attribute separation, so we do not report willingness to pay or simulated
choice shares.

## Sources

- [California Energy Commission ZEV sales](https://www.energy.ca.gov/data-reports/energy-almanac/zero-emission-vehicle-and-infrastructure-statistics-collection/new-zev)
- [U.S. Energy Information Administration gasoline prices](https://www.eia.gov/dnav/pet/pet_pri_gnd_dcus_sca_w.htm)
- [Google Trends](https://trends.google.com/trends/)
- [FHWA Traffic Volume Trends](https://www.fhwa.dot.gov/policyinformation/travel_monitoring/tvt.cfm)
